---
title: "How many people will be at my class reunion?"
author: "Victoria Leigh"
date: '`r format(Sys.time(), "%d %B %Y")`'
output: html_document
---

Rmd file to accompany the Shiny app.
This was used to test my calculations before I made the app and is slightly nicer to read in case I need to run a similar analysis again.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = F, message = F)
library(tidyverse) #for data wrangling 
library(forecast) 
```

###Default values 
```{r}
n_male <- 50 
n_female <- 50
age_grad <- 24
year_grad <- 1979
YoB_input <- year_grad - age_grad
current_year <- lubridate::year(Sys.Date())
current_age <- age_grad + (current_year - year_grad)
years_passed <- current_age - age_grad
```

###Input and tidy populations and deaths from (ONS, 2017) https://tinyurl.com/y7v5nvvf 
####Male 
```{r, message = F, warning = F}
#Populations
male_pop <- read_csv("male-pop.csv", col_types = cols(X58 = col_skip(), 
    X59 = col_skip(),
    X60 = col_skip(),
    X61 = col_skip(),
    X62 = col_skip(),
    X63 = col_skip(),
    X64 = col_skip()), 
    skip = 1) %>%
  head(-6) %>%
  gather(year, pop, 2:57)

#Deaths (better way of wranging the data)
male_deaths <- read_csv("male-deaths.csv", skip =1) %>%
  head(-3) %>%
  gather(year, deaths, 2:85) %>%
  na.omit()

#Join tables 
male <- inner_join(male_pop, male_deaths, by = c("age", "year")) %>%
  filter(age != "105+") %>%
  mutate(year = as.numeric(year),
         age = as.numeric(age),
         YoB = year- age) %>% #Create Year of Birth column
  filter(YoB == YoB_input, age>= age_grad) %>% #Keep ony the data from graduation onwards
  mutate(qt = deaths/ pop,  #probability of dying during the time interval 
         pt = 1 - qt,  #probability of surviving in the time interval 
         Psur = 1) #cumulative probability of surviving at the beginning of the time interval 

#Psur2 = pt * Psur1 starting from 100 survival
for(i in 2:nrow(male)) {
  male$Psur[i] <- male$pt[i - 1] * male$Psur[i - 1]
  }

```

####Female
```{r}
#The same as male 


female_pop <- read_csv("female-pop.csv", skip =1) %>%
  head(-6) %>%
  gather(year, pop, 2:62) %>%
  na.omit()

female_deaths <- read_csv("female-deaths.csv", skip =1) %>%
  head(-3) %>%
  gather(year, deaths, 2:78) %>%
  na.omit() %>%
  head(-11)

female <- inner_join(female_pop, female_deaths, by = c("age", "year")) %>%
  filter(age != "105+") %>%
  mutate(year = as.numeric(year),
         age = as.numeric(age),
         YoB = year- age) %>%
  filter(YoB == YoB_input, age >= age_grad) %>%
    mutate(qt = deaths/ pop, 
         pt = 1 - qt, 
         Psur = 1)

for(i in 2:nrow(female)) {
  female$Psur[i] <- female$pt[i - 1] * female$Psur[i - 1]
  }

```

###How many estimates need
```{r}
#The data only goes up to 2016 so to find the cumulative probabilty 
#for survival we need to forecast the data

last_age <- last(male$age)
last_year <- last(male$year)
n_missing <- current_age - last_age
```

###Forecast survival rates 
```{r}
#forecast function projects the rates forward
m_pred <- as_tibble(forecast(male$Psur, h = n_missing)) %>%
  mutate(Psur = `Point Forecast`,
          age = seq((last_age + 1), current_age, by = 1),
         year = seq((last_year +1), current_year, by = 1)) %>%
  select(age, year, Psur)

f_pred <- as_tibble(forecast(female$Psur, h = n_missing)) %>%
  mutate(Psur = `Point Forecast`,
          age = seq((last_age + 1), current_age, by = 1),
         year = seq((last_year +1), current_year, by = 1)) %>%
  select(age, year, Psur)

#select the current years probaility of survival and 
#mutiply by number of people in class

#Number of men who survived
m_pred_value <- m_pred$Psur[n_missing] * n_male

#Number of women who survived
f_pred_value <- f_pred$Psur[n_missing] * n_female 

#Total survived
total_pred <- round(m_pred_value + f_pred_value, 0)
```

`r total_pred` people from the class of `r year_grad` are expected 
to have survived to the age of `r current_age`.

###Create plot of survival rate
```{r}
m_to_join <- male %>%
  select(age, year, Psur)

f_to_join <- female %>%
  select(age, year, Psur)

#join together calculated probabilities to predicted probailities 
m_plot <- union(m_to_join, m_pred) %>%
  mutate(gender = "Male")

f_plot <- union(f_to_join, f_pred) %>%
  mutate(gender = "Female")

plot_data <- union(f_plot, m_plot)

thePlot <- ggplot(data = plot_data, aes(x = year, y= Psur * 100, colour = gender)) +
  geom_smooth() +
  scale_colour_manual("Gender", labels= c("Female", "Male"), values = c("gold", "darkgreen")) +
  theme_classic() +
  labs(title = "Percentage of expected survival since graduation", 
       x = "Year", y = "% Survival") +
  theme(legend.position = "bottom") +
  guides(color=guide_legend(override.aes=list(fill=NA))) 

thePlot
```

